\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}

### 2024-12-30 main.tex
update 2024 end

\section{Consideration}
\begin{itemize}
    \item AC equivalence and alpha equivalence.
    \item Efficient AC equivalence checking by sorting on syntax.
    \item Sorting requires that alpha equivalent terms have the same syntax.
    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
    \item only compute de Bruijn form in the equivalence checking and sorting phase.
\end{itemize}

\section{TODO}
\begin{itemize}
    \item implemented trace output for the whole pipeline.
    \item Note that the context should also be maintained during rewriting matching.
    \item better parser and output.
\end{itemize}

@@

\section{Language Syntax}
\begin{definition}[syntax]
    The syntax for type indices are defined as
    \begin{align*}
        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
    \end{align*}
    The syntax for Dirac notation types is defined as 
    \begin{align*}
        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
    \end{align*}
    The syntax for Dirac notation terms is defined as
    \begin{align*}
        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
    \end{align*}
\end{definition}
 % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent. For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types. To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

@@

\begin{figure}
    \begin{gather*}
        \textbf{Index-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
        \qquad \qquad
        \textbf{Index-Prod} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index}
    \end{gather*}
    \caption{Rules for type index.}
\end{figure}

@@

\begin{figure}[h]
    \begin{gather*}
        \textbf{Beta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda x : T.t)\ u)\ \reduce\ t\{x/u\}} 
        \qquad \qquad
        \textbf{Beta-Index} \qquad
        \frac{}{E[\Gamma] \vdash ((\mu x.t)\ u)\ \reduce\ t\{x/u\}} 
        \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
        \\
        \textbf{Eta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash \lambda x : T . (t\ x)\ \reduce\ t}
        \qquad \qquad
        \textbf{Eta-Index} \qquad
        \frac{}{E[\Gamma] \vdash \mu x . (t\ x)\ \reduce\ t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

@@

@@

>>>>

\section{Consideration}
\begin{itemize}
    \item AC equivalence and alpha equivalence.
    \item Efficient AC equivalence checking by sorting on syntax.
    \item Sorting requires that alpha equivalent terms have the same syntax.
    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
    \item only compute de Bruijn form in the equivalence checking and sorting phase.
    \item pipeline: 
        \begin{enumerate}
            \item preprocessing
            \item 1st rewritings
            \item variable expansion
            \item 2nd rewritings 
            \item computing de Bruijn and sorting
        \end{enumerate}

    \item I found that typing of inner bound variables also requires modifying the context.
    Need to modify the tree visiting algorithm
    (Typed term rewriting with bound variables is not that easy)    
\end{itemize}

\section{TODO}
\begin{itemize}
    \item implemented trace output for the whole pipeline.
    \item better parser and output.
    \item add Mathematica support for scalars
    \item output trace to Coq
    \item consider labelled Dirac notations
    \item better trace output
    \item better error logic
    \item better format output
\end{itemize}

\section{Things to Note}
\begin{itemize}
    \item The context should also be maintained during rewriting matching.
    \item We don't allow eta reduction. It will intertwine with SUM-SWAP and break the confluence.
\end{itemize}

@@

\section{Language Syntax}
\begin{definition}[syntax]
    The syntax for type indices are defined as
    \begin{align*}
        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \{c_0, c_1, \cdots, c_n\}.
    \end{align*}
    The syntax for Dirac notation types is defined as 
    \begin{align*}
        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
    \end{align*}
    The syntax for Dirac notation terms is defined as
    \begin{align*}
        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
    \end{align*}
\end{definition}
 % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent. For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types. To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

@@

\begin{figure}
    \begin{gather*}
        \textbf{Index-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
        \qquad \qquad
        \textbf{Index-Prod} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
        \\
        \textbf{Index-Concrete} \qquad
        \frac{\WF(E)[\Gamma] \qquad c_i \notin E[\Gamma] \text{ for all } i \qquad c_i \text{ are unique}}{E[\Gamma] \vdash \{c_1, c_2, \cdots, c_n\} : \Index}
    \end{gather*}
    \caption{Rules for type index.}
\end{figure}

@@

\begin{figure}[h]
    \begin{gather*}
        \textbf{Beta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda x : T.t)\ u)\ \reduce\ t\{x/u\}} 
        \qquad \qquad
        \textbf{Beta-Index} \qquad
        \frac{}{E[\Gamma] \vdash ((\mu x.t)\ u)\ \reduce\ t\{x/u\}} 
        \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

@@

\section{Preprocessing}
\begin{figure}
    \begin{gather*}
        \frac{E[\Gamma] \vdash B : \BType(\sigma) \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \circ K : \SType}
    \end{gather*}
    \caption{The typing rules for preprocessing}
\end{figure}

@@

Comment \texttt{(* ... *)} can be inserted between commands.

These are tye parsing rules for different expressions:
\begin{itemize}
    \item \texttt{Def ID := term.} --- \texttt{Def(ID term)}
    \item \texttt{Def ID := term : type.} --- \texttt{Def(ID term type)}
    \item \texttt{Var ID : type.} --- \texttt{Var(ID type)}
    \item \texttt{Check ID.} --- \texttt{Check(term)}
    \item \texttt{Show ID.} --- \texttt{Show(ID)}
    \item \texttt{ShowAll.} --- \texttt{ShowAll}
    \item \texttt{Normalize term.} --- \texttt{Normalize(term)}
    \item \texttt{Normalize term with trace.} --- \texttt{Normalize(term Trace)}
    \item \texttt{Check term = term.} --- \texttt{CheckEq(term term)}
    \item \texttt{T1 -> T2} --- \texttt{Arrow(T1 T2)}.
    \item \texttt{forall x. T} --- \texttt{Forall(x T)}
    \item \texttt{(e1, e2)} --- \texttt{PAIR(e1 e2)}
    \item \texttt{fun x : T => e} --- \texttt{fun(x T e)}
    \item \texttt{idx x => e} --- \texttt{idx(x e)}
    \item \texttt{e1 @ e2} --- \texttt{COMPO(e1 e2)}
    \item \texttt{e1 + ... + en} --- \texttt{ADDG(e1 ... en)}
    \item \texttt{e1 * ... * en} --- \texttt{STAR(e1 ... en)}
    \item \texttt{e1\^{}*} --- \texttt{CONJ(e1)}
    \item \texttt{delta(e1, e2)} --- \texttt{DELTA(e1 e2)}
    \item \texttt{|e>} --- \texttt{KET(e)}
    \item \texttt{<e|} --- \texttt{BRA(e)}
    \item \texttt{e1\^{}D} --- \texttt{ADJ(e1)}
    \item \texttt{e1.e2} --- \texttt{SCR(e1 e2)}
    \item \texttt{Sum(i in s, e)} --- \texttt{SSUM(i s e)}
\end{itemize}

These symbols will be transformed into internal language:
\begin{itemize}
    \item $A \circ B$ : $S \circ S$, $S \circ K$, $S \circ B$, $S \circ O$, $K \circ S$, $K \circ K$, $K \circ B$, $B \circ S$, $B \circ K$, $B \circ B$, $B \circ O$, $O \circ S$, $O \circ K$, $O \circ O$, $f\circ a$ (arrow), $f\circ a$ (index).
    \item \texttt{STAR(a ... b)} :  $\sigma_1 \times \sigma_2$, $\MULS(a \cdots b)$, $O_1 \otimes O_2$, $ M_1 \star M_2$
    \item \texttt{ADDG(e1 ... en)} : $\ADDS(e1 \cdots en)$, $\ADD(e1 \cdots en)$
    \item \texttt{SSUM(i S e)} : \texttt{SUM(s FUN(i T e))}
\end{itemize}
